---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

This assignment makes use of data from a personal activity monitoring
device. This device collects data at 5 minute intervals through out the
day. The data consists of two months of data from an anonymous
individual collected during the months of October and November, 2012
and include the number of steps taken in 5 minute intervals each day.

* Dataset: [Activity monitoring data](activity.zip) [52K]

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **interval**: Identifier for the 5-minute interval in which
    measurement was taken

We read **dataset** from zipped csv file and create additional variable

* **time**: Time of beginning of the 5-minute interval.

```{r}
dataset <- read.csv(unz("activity.zip", "activity.csv"), stringsAsFactors=FALSE)
dataset$time <- as.POSIXct(strptime(paste(dataset$date,
                                          dataset$interval %/% 100,
                                          dataset$interval %% 100),
                                    "%Y-%m-%d %H %M"))
dataset$date <- as.Date(dataset$date, "%Y-%m-%d")
```

We are using the fact that `interval` variable is just an integer converted
from string with `%H%M` format.

## What is mean total number of steps taken per day?

We will use dplyr package to manipulate data.

```{r}
library(dplyr, warn.conflicts = FALSE, quietly = TRUE)
```

First, convert dataset into datatable **dt** and produce new datatable by
grouping by date and summarizing by steps.

**by_day** contains 2 variables:

* **date**: The date on which the measurement was taken

* **total_steps**: Total number of steps for this day

```{r}
dt <- tbl_df(dataset)
by_day <- 
    dt %>%
    group_by(date) %>%
    summarize(total_steps = sum(steps, na.rm=TRUE))
```

```{r results="hide"}
max(by_day$total_steps)
```

Maximum amount of steps by day is `r max(by_day$total_steps)`, so we can create
a histogram with 24 bins.

```{r}
xbreaks <- seq(0, 24000, by = 1000)
hist(by_day$total_steps, breaks = xbreaks,
     xlab = "Total number of steps")
```

Compute mean total number of steps taken per day.

```{r}
mean(by_day$total_steps)
```

Compute median total number of steps taken per day.

```{r}
median(by_day$total_steps)
```

## What is the average daily activity pattern?

```{r}
by_time <-
    dt %>%
    group_by(interval) %>%
    summarize(avg_steps = mean(steps, na.rm=TRUE)) %>%
    mutate(hour = (interval %/% 100) + (interval %% 100) / 60)

plot(by_time$hour, by_time$avg_steps, type="l", xaxt="n")
axis(1, at=seq(0, 24, by = 2))

by_time %>%
    filter(avg_steps == max(avg_steps)) %>%
    print
```

## Imputing missing values

```{r}
sum(is.na(dt$steps))
```

```{r}
dt_full <- dt
for(i in 1:length(dt$steps)) {
    if (is.na(dt$steps[i])) {
        interval_num <- (dt$interval[i] %/% 100) * 12 +
            (dt$interval[i] %% 100) / 5 + 1
        dt_full$steps[i] <- by_time$avg_steps[interval_num]
    }    
}
```

```{r}
by_day_full <- 
    dt_full %>%
    group_by(date) %>%
    summarize(total_steps = sum(steps))

hist(by_day_full$total_steps, xbreaks)
mean(by_day_full$total_steps)
median(by_day_full$total_steps)
```

## Are there differences in activity patterns between weekdays and weekends?

```{r}
library(lubridate)
dt_full$week_day <- factor(ifelse(wday(dt_full$date) %in% c(1, 7), "weekend", "weekday"))
```

```{r}
by_weekdays <-
    dt_full %>%
    group_by(interval, week_day) %>%
    summarize(avg_steps = mean(steps))

library(ggplot2)
ggplot(by_weekdays, aes(interval, avg_steps)) +
    facet_wrap(~ week_day, nrow=2) + geom_line() +
    xlab("Interval") + ylab("Average number of steps")

ggplot(by_weekdays, aes((interval %/% 100) + (interval %% 100) / 60,
                        avg_steps)) + 
    facet_wrap(~ week_day, nrow=2) + geom_line() +
    xlab("Hour") + ylab("Average number of steps per 5-min interval")
```
